---
title: "Final Project"
author: "Brendan Gubbins"
date: "5/25/2021"
output:
  html_document:
    df_print: paged
  pdf_document:
    latex_engine: xelatex

---

## Loading the data

```{r}
rm(list=setdiff(ls(), c("lat", "lon")))

pacman::p_load(tidyverse, magrittr, data.table, R.utils)
housing = read.csv("C:\\Users\\Brendan Gubbins\\Desktop\\QC_MATH_342W_Spring_2021\\writing_assignments\\housing_data_2016_2017.csv")
housing = as_tibble(housing)
```

## Feature Selection

```{r}
housing_data = housing %>%
  select(approx_year_built, cats_allowed, common_charges, coop_condo, date_of_sale, dining_room_type, 
         dogs_allowed, fuel_type, full_address_or_zip_code, garage_exists, kitchen_type, 
         maintenance_cost, num_bedrooms, num_floors_in_building, num_full_bathrooms, num_half_bathrooms,
         parking_charges, sq_footage, total_taxes, sale_price, pct_tax_deductibl)

housing %>%
  filter(sq_footage > 6000 & !is.na(sale_price))

pacman::p_load(skimr)
skim(housing_data)
```

## Loading new data

```{r}
nyc_zip_codes = read.csv("C:\\Users\\Brendan Gubbins\\Downloads\\nyc-zip-codes.csv")

nyc_zip_codes %<>%
  select(ZipCode, Neighborhood) %>%
  rename(zip_code = ZipCode, neighborhood = Neighborhood)

# dropping an incomplete address
housing_data %<>%
  filter(full_address_or_zip_code != "11364")

housing_data %<>%
  mutate(zip_code = as.numeric(str_extract(substr(full_address_or_zip_code, 5, length(full_address_or_zip_code)), "[0-9]{5}")))

housing_data = left_join(housing_data, nyc_zip_codes, by = "zip_code")
```

## Converting `date_of_sale` into `day_of_sale`, `month_of_sale`. Also `approx_year_built` into `decade_built`

```{r}
pacman::p_load(lubridate)

housing_data %<>%
  mutate(day_of_sale = day(mdy(date_of_sale)),
         month_of_sale = month(mdy(date_of_sale))) %>%
  select(-date_of_sale)
```

## Converting `cats_allowed`, `dogs_allowed` into `pets_allowed` (binary), and `garage_exists` into binary

```{r}
housing_data %<>%
  mutate(cats_allowed = if_else(cats_allowed == "no", 0, 1)) %>%
  mutate(dogs_allowed = if_else(dogs_allowed == "no", 0, 1)) %>%
  mutate(garage_exists = if_else(is.na(garage_exists), 0, 1)) %>%
  mutate(pets_allowed = if_else(cats_allowed == 1 | dogs_allowed == 1, 1, 0)) %>%
  select(-cats_allowed, -dogs_allowed)
```

## Converting `coop_condo`, `dining_room_type`, `fuel_type`, `kitchen_type` into categorical features

```{r}
housing_data %<>%
  mutate(coop_condo = as.factor(coop_condo))

housing_data %<>%
  mutate(dining_room_type = if_else(dining_room_type == "none" | dining_room_type == "dining area", "other", dining_room_type)) %>%
  mutate(dining_room_type = as.factor(dining_room_type))

tabulate = sort(table(housing_data$fuel_type))

housing_data %<>%
  mutate(fuel_type = if_else(fuel_type %in% names(tabulate[tabulate < 62]), "other", fuel_type))

housing_data %<>%
  mutate(fuel_type = as.factor(fuel_type))

housing_data %<>%
  mutate(kitchen_type = if_else(kitchen_type == "eat in" | kitchen_type == "Eat in" | kitchen_type == "Eat In" | kitchen_type == "eatin", "eat in", kitchen_type)) %>%
  mutate(kitchen_type = if_else(kitchen_type == "combo" | kitchen_type == "Combo", "combo", kitchen_type)) %>%
  mutate(kitchen_type = if_else(kitchen_type == "efficiency" | kitchen_type == "efficiemcy" | kitchen_type == "efficiency kitchen"
                                | kitchen_type == "efficiency kitchene" | kitchen_type == "efficiency ktchen", "efficiency", kitchen_type))
housing_data %<>%
  mutate(kitchen_type = as.factor(kitchen_type))

housing_data = housing_data[housing_data$kitchen_type != "1955",]
```

## Cleaning `parking_charges`, `total_taxes`, `common_charges`, `sale_price`, `maintenance_cost`

```{r}
housing_data %<>%
  mutate(parking_charges = as.numeric(gsub("[\\$,]", "", parking_charges)),
         parking_charges = if_else(is.na(parking_charges), 0, parking_charges))

housing_data %<>%
  mutate(total_taxes = as.numeric(gsub("[\\$,]", "", total_taxes)))

# condos pay charges

housing_data %<>%
  mutate(common_charges = as.numeric(gsub("[\\$,]", "", common_charges)),
         common_charges = if_else(is.na(common_charges) & coop_condo == "co-op", 0, common_charges))

housing_data %<>%
  mutate(sale_price = as.numeric(gsub("[\\$,]", "", sale_price)))

housing_data = housing_data[!is.na(housing_data$sale_price),]

# co-ops pay maintenance 

housing_data %<>%
  mutate(maintenance_cost = as.numeric(gsub("[\\$,]", "", maintenance_cost)),
         maintenance_cost = if_else(is.na(maintenance_cost) & coop_condo == "condo", 0, maintenance_cost))
```

## `pct_tax_deductibl` applies to co-op only

```{r}
housing_data %<>%
  mutate(pct_tax_deductibl = if_else(housing_data$coop_condo == "condo", 0, as.numeric(pct_tax_deductibl)))
```

## Converting `NA` half bathrooms into `0`

```{r}
housing_data %<>%
  mutate(num_half_bathrooms = if_else(is.na(num_half_bathrooms), 0, as.numeric(num_half_bathrooms)))
```

## Converting `sq_footage` to percentiles

```{r}
# correcting an error
housing_data %<>%
  mutate(sq_footage = if_else(sq_footage > 6000, 1200, as.numeric(sq_footage)))

housing_data %<>%
  mutate(sq_footage = ecdf(sq_footage)(sq_footage))
```

## Geocoding

```{r}
latlon = geocode(housing_data$full_address_or_zip_code, output = "latlon")
lat = latlon$lat
lon = latlon$lon

housing_data %<>%
  mutate(latitude = lat,
         longitude = lon)

# grand central terminal
gc_coords = c(40.7527, 73.9772)
grand_central = array(NA, nrow(housing_data))

for (i in 1 : nrow(housing_data)) {
  grand_central[i] = distm(gc_coords, c(abs(housing_data$latitude[i]), abs(housing_data$longitude[i])), fun = distHaversine) / 1609
}

housing_data %<>%
  mutate(grand_central = grand_central)
```

## Missingness Dummy Variables

```{r}
M = tbl_df(apply(is.na(housing_data), 2, as.numeric))
colnames(M) = paste("is_missing_", colnames(housing_data), sep = "")
M %<>% 
  select_if(function(x){sum(x) > 0})

M = tbl_df(t(unique(t(M))))

housing_data %<>%
  relocate(sale_price)

housing_data = cbind(housing_data, M)
```

## Train-Test Split

```{r}
K = 5
test_prop = 1 / K
train_indices = sample(1 : nrow(housing_data), round((1 - test_prop) * nrow(housing_data)))
housing_train = housing_data[train_indices, ]
y_train = housing_train$sale_price
X_train = housing_train
X_train$sale_price = NULL
test_indices = setdiff(1 : nrow(housing_data), train_indices)
housing_test = housing_data[test_indices, ]
y_test = housing_test$sale_price
X_test = housing_test
X_test$sale_price = NULL
```

## Imputation with missForest

```{r}
pacman::p_load(missForest)

train_bind = cbind(X_train, sale_price = y_train)
test_bind = cbind(X_test, sale_price = NA)
X_bind = rbind(train_bind, test_bind)

X_bind$full_address_or_zip_code = NULL
neighborhood = X_bind$neighborhood
X_bind$neighborhood = NULL

Ximp = missForest(as.data.frame(X_bind))$ximp

# if coop, discount
Ximp %<>%
  mutate(monthly_cost = if_else(coop_condo == "co-op", 
                                (maintenance_cost + common_charges) * (100 - pct_tax_deductibl) / 100, 
                                maintenance_cost + common_charges))

Ximp %<>%
  select(-sale_price, -zip_code, -pct_tax_deductibl, -is_missing_pct_tax_deductibl)

Ximp = cbind(Ximp, neighborhood = as.factor(neighborhood), sale_price = X_bind$sale_price)

X_train = Ximp[!is.na(Ximp$sale_price),]
X_train$sale_price = NULL

X_test = Ximp[is.na(Ximp$sale_price),]
X_test$sale_price = NULL
```

## Regression Tree

```{r}
pacman::p_load(YARF)

tree_mod = YARFCART(X_train, y_train)

illustrate_trees(tree_mod, max_depth = 5, length_in_px_per_half_split = 25, open_file = TRUE)

tree_mod
```

## Linear Regression

```{r}
mod = lm(y_train ~ ., X_train)

summary(mod)$r.squared
sqrt(mean(mod$residuals^2))

y_hat = predict(mod, X_test)

residuals = y_test - y_hat
sqrt(mean(residuals^2))
```

## Random Forest

```{r}
rf_mod = YARF(X_train, y_train) 
rf_mod

y_hat = predict(rf_mod, X_test)

residuals = y_test - y_hat
oos_rmse = sqrt(mean(residuals^2))
oos_rmse
```